Improving \textsl{Gaia} parallaxes with a data-driven model \\ of the color--magnitude diagram

### 2017-03-16 paperDistance/paper.tex
notation

We would like to infer the distance to stars using the measured (noisy) parallax from Gaia. Gaia gives us the likelihood of the parallax of a star, given some unknown distance and uncertainties (assumed gaussian)
\begin{equation}
\label{eq:gaialike}
P(\varpi |d, \sigma^2_{\varpi}, \textbf{I})
\end{equation}

where $d$ is distance to the star, $\varpi$ is the observed parallax, and $\sigma^2_{\varpi}$ is the uncertainty in that measurement, and I are other various prior assumptions (mostly about the noise properties?). This is the probability of the data given the model distance, but we want to flip it around and infer the distance to the star given the measured parallax and it's uncertainty. We do this using probability theory/bayes rule, and turn the likelihood of the data into the posterior probability of the distance

\begin{equation}
\label{eq:bayes}
P(d_{true}|\varpi, \sigma^2_{\varpi}, \textbf{I}) = \frac{P(\varpi |d_{true} \sigma^2_{\varpi}, \textbf{I}) P(d)}{\textbf{Z}},
\end{equation}

where $P(d)$ is our prior knowledge of the distance to the star, and $\textbf{Z}$ is a normalization constant to make the posterior sum to 1. The strength in our inference method lies in how we generate our prior knowledge of the distances to stars. Here, instead of assuming the spatial distribution of stars in the milky way, or relying on stellar models, as previous work as shown, we build the color-magnitude diagram of the observed stars from their observations as our prior. The absolute magnitude of the CMD carries information about the distances to stars and where they can physically lie in this space. Toy model?

@@

True values:
  C - true color,
  M - true absolute magnitude
  Y = [10.^0.2(M+2), C]
data:
    (J-K)_c
    \varpi
    J_c
    y = [\varpi10^0.2J_c, (J-K)_c]

  prior assumptions:
    \sigma_J, sigma_K, sigma_\varpi, Greene map

@@

\begin{equation}
\label{eq:data}
\begin{aligned}
\mathrm{true vector} \;\;\;\; &\vec{Y} = [10^{0.2\,M + 2}, C] \\
\mathrm{data vector} \;\;\;\; &\vec{y} = [\varpi\,10^{0.2J_{c,n}}, (J- K)_{c,n}] \;  n \in \mathrm{stars} \\
\mathrm{with \; absolute \; magnitude} \;\;\;\; &M_{J,n} = J_n - DM_n - Q_JE(B-V)_n \\
& DM_n = -5\log_{10}\frac{1}{\varpi} \\
\mathrm{and \; color} \;\;\;\; &(J - K)_n = J_n - K_n - Q_{JK}E(B-V)_n \\
\mathrm{assumptions} \;\;\;\; &I = \left\{\left\{\sigma^2_n\right\}_{n=1}^N, \mathrm{assumptions}\right\}
\end{aligned}
\end{equation}

@@

Likelihood function:
  P(y|Y) = N(y|Y,I)
Working with these variables, we then have a likelihood for the data, both the Gaia parallaxes and 2mass photometry of

\begin{equation}
\label{eq:fulllike}
P(\vec{y}|\vec{Y}) = N(\vec{y}|\vec{Y}, I)
\end{equation}

@@

\begin{equation}
\label{eq:xdmml}
\begin{aligned}

\alpha =& \left\{\mathrm{amp}_k, \mu_k, V_k\right\}_{k=1}^K \; k \in \mathrm{Gaussians} \\
       =& \argmax_{\alpha} \Pi_n P(\vec{y}_n|\alpha) \\
       =& \argmax_{\alpha} \Pi_n \int P(\vec{y}_n|\vec{Y}_n,I)P(\vec{Y}|\alpha)d\vec{Y}
\end{aligned}
\end{equation}

\begin{equation}
\label{eq:posterior}
P(Y|y, \alpha, I) = 1/Z p(y| Y,I)p(Y|\alpha, I)
\end{equation}

\begin{equation}
\label{eq:parallaxPost}
P(\vec{Y}_1|\vec{y}, \alpha, I) => p(\varpi_{true}|y, \alpha, I) \, f(\varpi_{true})
\[
f(\varpi_{true}) = \begin{cases}
              1: -2 < log_{10} \varpi < 2\\
              0: otherwise
              \end{cases}
\]
\end{equation}

>>>>

We would like to infer the distance to stars using the measured (noisy) parallax from Gaia. Gaia gives us the likelihood of the parallax of a star, given some unknown distance and uncertainties (assumed gaussian)
\begin{equation}
\label{eq:gaialike}
P(\varpi |\varpi_{true}, \sigma^2_{\varpi}, \textbf{I})
\end{equation}

where $d$ is distance to the star, $\varpi$ is the observed parallax, and $\sigma^2_{\varpi}$ is the uncertainty in that measurement, and I are other various prior assumptions (mostly about the noise properties which we will drop in our notation but is implicitly there). This is the probability of the data given the model distance, but we want to flip it around and infer the distance to the star given the measured parallax and it's uncertainty. We do this using probability theory/bayes rule, and turn the likelihood of the data into the posterior probability of the distance

\begin{equation}
\label{eq:bayes}
P(\varpi_{true}|\varpi, \sigma^2_{\varpi}) = \frac{P(\varpi |\varpi_{true}, \sigma^2_{\varpi}) P(\varpi_{true})}{\textbf{Z}},
\end{equation}

where $P(\varpi_{true})$ is our prior knowledge of the distance to the star, and $\textbf{Z}$ is a normalization constant to make the posterior sum to 1. The strength in our inference method lies in how we generate our prior knowledge of the distances to stars. Here, instead of assuming the spatial distribution of stars in the milky way, or relying on stellar models, as previous work as shown, we build the color-magnitude diagram of the observed stars from their observations as our prior. The absolute magnitude of the CMD carries information about the distances to stars and where they can physically lie in this space. Toy model?

@@

@@

\begin{equation}
\label{eq:data}
\begin{aligned}
\mathrm{true \, vector} \;\;\;\; &\vec{Y} = [10^{0.2\,M_{J,n} + 2}, C_n] \\
\mathrm{data \, vector} \;\;\;\; &\vec{y} = [\varpi\,10^{0.2J^c_n}, (J- K)^c_n] \;  n \in \mathrm{stars} \\
\mathrm{with \; absolute \; magnitude} \;\;\;\; &M^c_{J,n} = J_n + 5\log_{10}\varpi_n - Q_JE(B-V)_n \\
\mathrm{and \; color} \;\;\;\; &(J - K)^c_n = J_n - K_n - Q_{JK}E(B-V)_n \\
\mathrm{assumptions} \;\;\;\; &I = \left\{\left\{\sigma^2_{\varpi, n}\right\}, \left\{\sigma^2_{J,n}\right\}, \left\{\sigma^2_{K,n}\right\}, \mathrm{Dust \, Map}\right\}
\end{aligned}
\end{equation}

@@

Working with these variables, we then have a likelihood for the data, both the Gaia parallaxes and 2mass photometry of

\begin{equation}
\label{eq:fulllike}
P(\vec{y}|\vec{Y}) = \mathcal{N}(\vec{y}|\vec{Y}, \sigma_{\vec{y}})
\end{equation}

@@

\begin{equation}
\label{eq:xdmml}
\begin{split}
\alpha =& \left\{\mathrm{amp}_k, \mu_k, V_k\right\}_{k=1}^K \; k \in \mathrm{Gaussians} \\
       =& \argmax_{\alpha} \, \Pi_n \, P(\vec{y}_n|\alpha) \\
       =& \argmax_{\alpha} \, \Pi_n \, \int P(\vec{y}_n|\vec{Y}_n) \, P(\vec{Y}|\alpha)d\vec{Y} \\
\end{split}
\end{equation}

\begin{equation}
\label{eq:posterior}
P(\vec{Y}|\vec{y}, \alpha) = \frac{1}{Z} \, P(\vec{y}|\vec{Y}) \, P(\vec{Y}|\alpha)
\end{equation}

\begin{equation}
\label{eq:parallaxPost}
\begin{split}
P(\vec{Y}_0|\vec{y}, \alpha) => P(\varpi_{true}|y, \alpha) \, f(\varpi_{true}) \\
\[
f(\varpi_{true}) = \begin{cases}
              1: -2 < \mathrm{log}_{10} \varpi_{true} < 2\\
              0: \mathrm{otherwise}
              \end{cases}
\]
\end{split}
\end{equation}
